Position: Inflated AGI Claims Require Strong Tests and Dovelation \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-05-27 body.tex
fixed typo
outlook section added related works section almost done

To strengthen our argument, in this paper, we first consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made. We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made. Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there. Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.

@@

In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}}
%(formerly Twitter)
platform~\citep{gurnee2023languagev1}. Through linear probing experiments, the claim was made that LLMs learned literal maps of the world. As such, they were considered to be more than `stochastic parrots'~\citep{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it. While the manuscript immediately received public criticism~\citep{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\citep{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings. However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation. After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.

@@

\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}

@@

While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, %with this simple example, we
this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features. Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification. In fact, it is well-known that random projections can be used for prediction tasks \citep{dasgupta2013experiments}.

@@

\section{Human Proneness to Over-Interpretation}

@@

\section{Conclusion}
As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests. Even academic researchers are not free from such biases. Thus, we call for the community to create explicit room for organized skepticism.

>>>>

To strengthen our argument, in this paper, we first present related work in Section~\ref*{related}. We then consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made. In Section~\ref*{patterns-in-latent-spaces-and-how-to-find-them}, we present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made. Furthermore, we present a review of social science findings in Section~\ref*{social} that underline how prone humans are to being enticed by patterns that are not really there. Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this. Finally, in Section~\ref*{outlook}, we propose specific structural and cultural changes to improve the current situation. Section~\ref*{conclude} concludes.

\section{Related Work}\label{related}

In this work, we question the practice of using outcomes from mechanistic interpretability to support AGI claims. This is not to be seen as criticism toward the underlying methodologies in isolation, but rather in the context of current publishing practices that we repeatedly challenge throughout this work. Many closely related works are free of any grandiose conclusions and instead highlight the benefits of mechanistic interpretability that we also highlight in this work~\citep{nanda2023emergent,gurnee2023finding,li2022emergent}.

Another related subfield investigates the capacity of LLMs to reason causally. Here, too, there is an opportunity to over-interpret the finding of causal information as causal understanding. Recent work has shown, for example, that LLMs can indeed correctly predict causal relationships and this may have practical use cases~\citep{kiciman2023causal}. But despite the potential utility, the authors also demonstrate that this capacity can be partially explained by memorization, rather than an actual understanding of causal relationships. Similarly,~\citet{zevcevic2023causal} provide empirical evidence indicating that current LLMs ``may talk causality but are not causal''.

Finally, another recent work closely aligned with the position we present here is~\ref*{schaeffer2024emergent}.

@@

In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} platform~\citep{gurnee2023languagev1}. Through linear probing experiments, the claim was made that LLMs learned literal maps of the world. As such, they were considered to be more than `stochastic parrots'~\citep{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it. While the manuscript immediately received public criticism~\citep{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\citep{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings. However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation. After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.

@@

\caption{\label{fig-map}Predicted coordinate values (out-of-sample) from a linear probe on final-layer activations of an untrained neural network.}

@@

While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features. Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification. In fact, it is well-known that random projections can be used for prediction tasks \citep{dasgupta2013experiments}.

@@

\section{Human Proneness to Over-Interpretation}\label{social}

@@

\section{Outlook}\label{outlook}

We believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry. Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship. This argument has been raised long before both in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}. Specifically,~\citet{liem2023treat} argue that societally impactful scientific insights should be treated as open-source software artifacts. The open-source community sets a positive example of how scientific artifacts should be published in many different ways. Not only does it adequately reward small contributions but it also naturally considers negative results (bugs) as part of the scientific process. Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects. Open review platforms like OpenReview are a step in the right direction but they still fall short of what is technically feasible. Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged. As we have pointed out repeatedly in this work, any claims about AGI demand proper strong hypothesis tests. It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making claims.

\section{Conclusion}\label{conclude}
As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests. Even academic researchers are not free from such biases. Thus, we call for the community to create explicit room for organized skepticism.
